String Constraints with Regex-Counting and String-Length Solved Efficiently

### 2023-05-05 2023-1-19/sections/algorithms.tex
solving NE(CEFA)

\begin{definition}[Nonemptiness of CEFA w.r.t. an LIA formula\footnote{The decision problem was called satisfiability of an LIA formula w.r.t. CEFA in \cite{atva2020}.}, abbreviated as $\cefadec$]
Suppose that $x_1, \cdots, x_n$ are mutually distinct string variables, $\Lambda_{x_1}, \cdots, \Lambda_{x_n}$ are nonempty sets of CEFA over the same alphabet $\Sigma$ with $\Lambda_{x_i} = \{\aut_{i, 1}, \cdots, \aut_{i, l_i}\}$ for every $i \in [n]$, moreover, the sets of registers $R_{\aut_{1, 1}}, \cdots, R_{\aut_{1, l_1}}, \cdots, R_{\aut_{n, 1}}, \cdots, R_{\aut_{n, l_n}}$ are mutually disjoint, and $\varphi$ is an LIA formula whose free variables are from $R' = \bigcup \limits_{i \in [n], j \in [l_i]} R_{\aut_{i, j}}$. Then  the CEFA in $\Lambda_{x_1}, \cdots, \Lambda_{x_n}$ are said to be nonempty w.r.t. $\varphi$ if there are an assignment $\eta: \{x_1, \cdots, x_n\} \rightarrow \Sigma^*$ and vectors 
$\myvec{v_{1,1}}, \cdots, \myvec{v_{1, l_1}}, \cdots, \myvec{v_{n, 1}}, \cdots, \myvec{v_{n, l_n}}$ such that $(w_i; \myvec{v_{i,j}}) \in \Lang(\aut_{i, j})$ for every $i \in [n], j \in [l_i]$,
% \in \aut_{i,1}(\eta(x_1)), \cdots, \myvec{v_{i, l_i}} \in \aut_{i,l_i}(\eta(x_i))$ for every $i \in [n]$  
%\cdots, \myvec{v_{n,1}} \in \aut_{n,1}(w_n), \cdots, \myvec{v_{n, l_n}} \in \aut_{n, l_n}(w_n)$ 
and $\varphi[\myvec{v_{1,1}}/R_{\aut_{1,1}}, \cdots, \myvec{v_{n, l_n}}/R_{\aut_{n, l_n}}]$ is true. 
\end{definition}

@@

In the sequel, we present a procedure to solve the $\cefadec$ problem.

Suppose that $x_1, \cdots, x_n$ are mutually distinct string variables, $\Lambda_{x_1}, \cdots, \Lambda_{x_n}$ are nonempty sets of CEFA over the same alphabet $\Sigma$ with $\Lambda_{x_i} = \{\aut_{i, 1}, \cdots, \aut_{i, l_i}\}$ for every $i \in [n]$, moreover, the sets of registers $R_{\aut_{1, 1}}, \cdots, R_{\aut_{1, l_1}}, \cdots, R_{\aut_{n, 1}}, \cdots, R_{\aut_{n, l_n}}$ are mutually disjoint, and $\varphi$ is an LIA formula whose free variables are from $R' = \bigcup \limits_{i \in [n], j \in [l_i]} R_{\aut_{i, j}}$.

At first, we compute CEFA $\cB_{i} = \aut_{i,1} \cap \cdots \cap \aut_{i, l_i}$ for each $i \in [n]$. For each $i \in [n]$, let $\Lambda'_{x_i} = \{\cB_i\}$. Evidently, $R_{\cB_i}= \bigcup \limits_{j \in [l_i]} R_{\aut_{i, j}}$ for each $i \in [n]$.
Then the nonemptiness of CEFA in $\Lambda_{x_1}, \cdots, \Lambda_{x_n}$ w.r.t. $\varphi$ is reduced to the nonemptiness of CEFA in $\Lambda'_{x_1}, \cdots, \Lambda'_{x_n}$ w.r.t. $\varphi$.

Since the construction of $\cB_1, \cdots, \cB_n$ may involve exponential blow-up, it is vital to reduce the sizes of these CEFAs.

It turns out that to solve the nonemptiness of CEFA in $\Lambda'_{x_1}, \cdots, \Lambda'_{x_n}$ w.r.t. $\varphi$, we can ignore the characters in $\cB_1,\cdots, \cB_n$. This is due to the fact that there is only one CEFA for each string variable and $\varphi$ only cares about the values of registers.

@@

>>>>

\begin{definition}[Nonemptiness of CEFA w.r.t. an LIA formula\footnote{The decision problem was called satisfiability of an LIA formula w.r.t. CEFA in \cite{atva2020}.}, abbreviated as $\cefadec$]
Suppose that $x_1, \cdots, x_n$ are mutually distinct string variables, $\Lambda_{x_1}, \cdots, \Lambda_{x_n}$ are nonempty sets of CEFA over the same alphabet $\Sigma$ with $\Lambda_{x_i} = \{\aut_{i, 1}, \cdots, \aut_{i, l_i}\}$ for every $i \in [n]$, moreover, the sets of registers $R_{\aut_{1, 1}}, \cdots, R_{\aut_{1, l_1}}, \cdots, R_{\aut_{n, 1}}, \cdots, R_{\aut_{n, l_n}}$ are mutually disjoint, and $\varphi$ is an LIA formula whose free variables are from $R' = \bigcup \limits_{i \in [n], j \in [l_i]} R_{\aut_{i, j}}$. Then  the CEFA in $\Lambda_{x_1}, \cdots, \Lambda_{x_n}$ are said to be nonempty w.r.t. $\varphi$ if there are an assignment $\theta: \{x_1, \cdots, x_n\} \rightarrow \Sigma^*$ and vectors 
$\myvec{v_{1,1}}, \cdots, \myvec{v_{1, l_1}}, \cdots, \myvec{v_{n, 1}}, \cdots, \myvec{v_{n, l_n}}$ such that $(\theta(x_i); \myvec{v_{i,j}}) \in \Lang(\aut_{i, j})$ for every $i \in [n], j \in [l_i]$,
% \in \aut_{i,1}(\eta(x_1)), \cdots, \myvec{v_{i, l_i}} \in \aut_{i,l_i}(\eta(x_i))$ for every $i \in [n]$  
%\cdots, \myvec{v_{n,1}} \in \aut_{n,1}(w_n), \cdots, \myvec{v_{n, l_n}} \in \aut_{n, l_n}(w_n)$ 
and $\varphi[\myvec{v_{1,1}}/R_{\aut_{1,1}}, \cdots, \myvec{v_{n, l_n}}/R_{\aut_{n, l_n}}]$ is true. 
\end{definition}

@@

In the sequel, we present a procedure to solve the $\cefadec$ problem: Suppose that $x_1, \cdots, x_n$ are mutually distinct string variables, $\Lambda_{x_1}, \cdots, \Lambda_{x_n}$ are nonempty sets of CEFA over the same alphabet $\Sigma$ with $\Lambda_{x_i} = \{\aut_{i, 1}, \cdots, \aut_{i, l_i}\}$ for every $i \in [n]$, moreover, the sets of registers $R_{\aut_{1, 1}}, \cdots, R_{\aut_{n, l_n}}$ are mutually disjoint, and $\varphi$ is an LIA formula whose free variables are from $R' = \bigcup \limits_{i \in [n], j \in [l_i]} R_{\aut_{i, j}}$.

The procedure comprises four steps. 
%
\paragraph*{Step 1.} For each $i \in [n]$, compute a CEFA $\cB_{i} = \aut_{i,1} \cap \cdots \cap \aut_{i, l_i}$, and let $\Lambda'_{x_i} := \{\cB_i\}$. \qed

%At first, we compute CEFA $\cB_{i} = \aut_{i,1} \cap \cdots \cap \aut_{i, l_i}$ for each $i \in [n]$. For each $i \in [n]$, let $\Lambda'_{x_i} = \{\cB_i\}$. Evidently, $R_{\cB_i}= \bigcup \limits_{j \in [l_i]} R_{\aut_{i, j}}$ for each $i \in [n]$.
After Step 1, the nonemptiness of CEFA in $\Lambda_{x_1}, \cdots, \Lambda_{x_n}$ w.r.t. $\varphi$ is reduced to the nonemptiness of CEFA in $\Lambda'_{x_1}, \cdots, \Lambda'_{x_n}$ w.r.t. $\varphi$.

%CEFA in $\Lambda'_{x_1} = \{\cB_1\}, \cdots, \Lambda'_{x_n} = \{\cB_n\}$ are nonempty w.r.t. $\varphi$ 
%

%Therefore, to solve the nonemptiness of CEFA in $\Lambda'_{x_1} = \{\cB_1\}, \cdots, \Lambda'_{x_n} = \{\cB_n\}$ w.r.t. $\varphi$, it is sufficient to compute $\cefaout(\cB_1)$

%Since the construction of $\cB_1, \cdots, \cB_n$ involves the computation of the product of multiple CEFA, it incurs an exponential blow-up in the worst case. Therefore, it is vital to reduce the sizes of these CEFAs. In the sequel, we propose some techniques to reduce the automata sizes.

In the sequel, we reduce the nonemptiness of CEFA in $\Lambda'_{x_1}, \cdots, \Lambda'_{x_n}$ w.r.t. $\varphi$ to the satisfiability of an LIA formula. The reduction relies on two observations.

The first observation follows from the fact that $\Lambda'_{x_i}$ for each $i \in [n]$ contains exactly one CEFA.

\begin{observation}\label{obs-cefa-output}
CEFA in $\Lambda'_{x_1} = \{\cB_1\}, \cdots, \Lambda'_{x_n} = \{\cB_n\}$ are nonempty w.r.t. $\varphi$ iff there are $\myvec{v_{1}} \in \cefaout(\cB_1), \cdots, \myvec{v_{n}} \in \cefaout(\cB_n)$ such that $\varphi[\myvec{v_{1}}/R_{\cB_{1}}, \cdots, \myvec{v_{n}}/R_{\cB_{n}}]$ is true.
\end{observation}

%From the fact that $\Lambda'_{x_i}$ for each $i \in [n]$ contains exactly one CEFA, we deduce that the following two conditions are equivalent.  
%\begin{itemize}
%\item CEFA in $\Lambda'_{x_1} = \{\cB_1\}, \cdots, \Lambda'_{x_n} = \{\cB_n\}$ are nonempty w.r.t. $\varphi$, namely, 
%there are an assignment $\theta: \{x_1, \cdots, x_n\} \rightarrow \Sigma^*$ and vectors $\myvec{v_{1}}, \cdots, \myvec{v_{n}}$ such that $(\theta(x_1); \myvec{v_1}) \in \Lang(\cB_1)$, $\cdots$, $(\theta(x_n); \myvec{v_n}) \in \Lang(\cB_n)$ and $\varphi[\myvec{v_{1}}/R_{\cB_{1}}, \cdots, \myvec{v_{n}}/R_{\cB_{n}}]$ is true,
%
%\item there are $\myvec{v_{1}} \in \cefaout(\cB_1), \cdots, \myvec{v_{n}} \in \cefaout(\cB_n)$ such that $\varphi[\myvec{v_{1}}/R_{\cB_{1}}, \cdots, \myvec{v_{n}}/R_{\cB_{n}}]$ is true.
%\end{itemize}

Let $\aut = (R, Q, \Sigma, \delta, I, F)$ be a CEFA and $\costset_\aut = \{\myvec{v} \mid \exists q, a, q'.\ (q, a, q', \myvec{v}) \in \delta\}$. 
Moreover, let $\aut'$ be an NFA over the alphabet $\costset_\aut$ that is obtained from $\aut$ by ignoring the characters, that is, the transitions of $\NFA'$ are the tuples $(q, \myvec{v}, q')$ such that $(q, a, q', \myvec{v}) \in \delta$ for some $a \in \Sigma$.
%Then $\cefaout(\aut) = \parikh(\NFA')$.
\begin{observation}\label{obs-cefa-nfa}
For each CEFA $\aut$, $\cefaout(\aut) = \left\{\sum \limits_{\myvec{v} \in \costset_\aut} \eta(\myvec{v}) \myvec{v}  \mid \eta \in \parikh(\NFA') \right\}$.
\end{observation}

From Observation~\ref{obs-cefa-nfa}, we know that the following two conditions are equivalent, 
\begin{itemize}
%\item CEFA in $\Lambda'_{x_1} = \{\cB_1\}, \cdots, \Lambda'_{x_n} = \{\cB_n\}$ are nonempty w.r.t. $\varphi$, namely, 
%there are an assignment $\theta: \{x_1, \cdots, x_n\} \rightarrow \Sigma^*$ and vectors $\myvec{v_{1}}, \cdots, \myvec{v_{n}}$ such that $(\theta(x_1); \myvec{v_1}) \in \Lang(\cB_1)$, $\cdots$, $(\theta(x_n); \myvec{v_n}) \in \Lang(\cB_n)$ and $\varphi[\myvec{v_{1}}/R_{\cB_{1}}, \cdots, \myvec{v_{n}}/R_{\cB_{n}}]$ is true,
%
\item there are $\myvec{v_{1}} \in \cefaout(\cB_1), \cdots, \myvec{v_{n}} \in \cefaout(\cB_n)$ such that $\varphi[\myvec{v_{1}}/R_{\cB_{1}}, \cdots, \myvec{v_{n}}/R_{\cB_{n}}]$ is true, 
%CEFA in $\Lambda'_{x_1} = \{\cB_1\}, \cdots, \Lambda'_{x_n} = \{\cB_n\}$ are nonempty w.r.t. $\varphi$ 
%
\item 
there are $\eta_1 \in \parikh(\cB'_1), \cdots, \eta_n \in \parikh(\cB'_n)$ such that 
%
%\begin{equation}\label{eqn-cefa-lia}
\[
\varphi \left[\sum \limits_{\myvec{v} \in \costset_{\cB_1}} \eta(\myvec{v}) \myvec{v} \big{/}R_{\cB_1}, \cdots, \sum \limits_{\myvec{v} \in \costset_{\cB_n}} \eta(\myvec{v}) \myvec{v} \big{/}R_{\cB_n} \right]
\]
%\end{equation} 
%
is true. 
\end{itemize}

Therefore, to solve the nonemptiness of CEFA in $\Lambda'_{x_1}, \cdots, \Lambda'_{x_n}$ w.r.t. $\varphi$, it is sufficient to compute the existential LIA formulas $\psi_{\cB'_1}(\fZ_{\costset_{\cB_1}}), \cdots, \psi_{\cB'_n}(\fZ_{\costset_{\cB_n}})$, which define the Parikh images of $\cB'_1$, $\cdots$, $\cB'_n$ respectively, and solve the satisfiability of the following existential LIA formula
\begin{equation}\label{eqn-cefa-lia}
\bigwedge \limits_{i \in [n]} \psi_{\cB'_i}(\fZ_{\costset_{\cB_i}}) \wedge 
\varphi \left[ \sum\limits_{\myvec{v} \in \costset_{\cB_1}} \anivarz_{\myvec{v}} \myvec{v} \big{/}R_{\cB_1}, \cdots, \sum\limits_{\myvec{v} \in \costset_{\cB_n}} \anivarz_{\myvec{v}} \myvec{v} \big{/}R_{\cB_n} \right],
\end{equation}
where $\anivarz_{\myvec{v}}$'s are integer variables to represent the number of occurrences of $\myvec{v}$.

Because the sizes of the LIA formulas $\psi_{\cB'_1}(\fZ_{\costset_{\cB_1}}), \cdots, \psi_{\cB'_n}(\fZ_{\costset_{\cB_n}})$ are proportional to the sizes (more precisely, the alphabet size, the number of states and transitions) of NFA $\cB'_1, \cdots, \cB'_n$, and it is expensive (NP-complete) to solve the satisfiability of existential LIA formulas, it is vital to reduce the sizes of these NFA.

We also have the following observation: For each $i \in [n]$ and $\eta \in \parikh(\cB'_i)$, we have 
$$\sum \limits_{\myvec{v} \in \costset(\cB_i)} \eta(\myvec{v}) \myvec{v} = \sum \limits_{\myvec{v} \in \costset(\cB_i) \setminus \{\myvec{0}\}} \eta(\myvec{v}) \myvec{v}.$$
In other words, the $\myvec{0}$-labeled transitions in $\cB'_i$ do not contribute to the sum $\sum \limits_{\myvec{v} \in \costset(\cB_i)} \eta(\myvec{v}) \myvec{v}$.

Therefore, we can reduce the sizes of $\cB'_1, \cdots, \cB'_n$ by Step II presented in the sequel.

\paragraph*{Step 2.} For each $i \in [n]$, view the transitions $(q, \myvec{0}, q')$ in $\cB'_i$ as $\varepsilon$-transitions $(q, \varepsilon, q')$, then remove the $\varepsilon$-transitions from $\cB'_i$, determinize and minimize the resulting NFA.  \qed

For $i \in [n]$, let  $\cC_i$ denote the DFA obtained from $\cB'_i$ by executing Step 2. From the construction, we know that $\parikh(\cC_i) = \parikh(\cB'_i)$ for each $i \in [n]$.
Therefore, we compute LIA formulas from $\cC_i$'s, instead of $\cB'_i$'s, to represent Parikh images.

\paragraph*{Step 3.} For each $i \in [n]$, compute an existential LIA formula $\psi_{\cC_i}(\fZ_{\costset_{\cB_i}})$ from $\cC_i$ to represent $\parikh(\cC_i)$. Then solve the satisfiability of the following existential LIA formula 
%
\[
\bigwedge \limits_{i \in [n]} \psi_{\cC_i}(\fZ_{\costset_{\cB_i}}) \wedge 
\varphi \left[ \sum\limits_{\myvec{v} \in \costset_{\cB_1}} \anivarz_{\myvec{v}} \myvec{v} \big{/}R_{\cB_1}, \cdots, \sum\limits_{\myvec{v} \in \costset_{\cB_n}} \anivarz_{\myvec{v}} \myvec{v} \big{/}R_{\cB_n} \right].
\]

Finally, although we propose techniques to reduce the sizes of $\cB'_i$'s, it is still relatively expensive to compute LIA formulas from NFA and solve their satisfiability, especially for those problem instances with positive answers, namely, those instances where CEFA in $\Lambda_{x_1}, \cdots, \Lambda_{x_n}$ are nonempty w.r.t. $\varphi$.   
Therefore, in the sequel, we propose a heuristic to search for solutions for those positive problem instances by utilizing under approximations, before Step 3. Intuitively, we search for solutions by bounding the lengths of the strings that are accepted by $\cC_i$'s. The heuristic uses a constant bound $B > 0$. Since the heuristic is applied before Step 3, we call it Step 2.5, to denote the fact that it is applied after Step 2 and before Step 3.

For $i \in [n]$, suppose $R_{\cB_i} = (r_{i,1},\cdots, r_{i, k_i})$ and $\cC_i = (Q'_i, \costset_{\cB_i}, \delta'_i, q_{i,0}, F'_i)$.

\paragraph*{Step 2.5.} Initially, let $len := 0$, and for each $i \in [n]$, let $\confset_{\cC_i}:=\{(q_{i,0}, \myvec{0})\}$ and $\psi_{\cC_i}:= \lfalse$.  Then iterate the following procedure until $len = B$.
\begin{quote}
For each $i \in [n]$, let $\confset'_{\cC_i}:=\confset_{\cC_i}$ and do the following:  For each $(q, \myvec{v}) \in \confset_{\cC_i}$,
\begin{itemize}
\item if $q \in F'_i$, then let $\psi_{\cC_i}:=\psi_{\cC_i} \vee (\bigwedge \limits_{j \in [k_i]} r_{i,j} = v_j)$, 
%
\item let $\confset'_{\cC_i}:=\confset'_{\cC_i} \setminus \{(q, \myvec{v})\}$, 
%
\item for each transition $(q, \myvec{v'}, q') \in \delta'_i$, let $\confset'_{\cC_i}:=\confset'_{\cC_i} \cup \{(q', \myvec{v}+ \myvec{v'})\}$. 
\end{itemize}
Let $\varphi:=\varphi \wedge \psi_{\cC_i}$.

If $\varphi$ is satisfiable, then return $\ltrue$. Let $len: = len+1$. \qed
\end{quote}

%Furthermore, for each $i \in [n]$, we can remove the redundancy of characters/vectors in $\cC_i$.

%%%%%%%%%%%% the original texts by Denghang %%%%%%%%%
%%%%%%%%%%%% the original texts by Denghang %%%%%%%%%
\hide{
Moreover, the transitions $(q, \myvec{0}, q')$ in $\cC_i$ can be seen as $\varepsilon$-transitions, that is, we can replace each transition $(q, \myvec{0}, q')$ with $(q, \varepsilon, q')$, without affecting the set $\{ \sum\limits_{\myvec{v} \in V_i} m_{\myvec{v}} \myvec{v} \mid \myvec{m} \in parikh(\cC_i)\}$. Therefore, we first do these replacements, apply the $\varepsilon$-transition removal procedure. Then we utilize the determinization and minimization algorithm for NFA to reduce the size of $\cC_i$.

It turns out that to solve the nonemptiness of CEFA in $\Lambda'_{x_1} = \{\cB_1\}, \cdots, \Lambda'_{x_n} = \{\cB_n\}$ w.r.t. $\varphi$, we can ignore the characters in $\cB_1,\cdots, \cB_n$. This is due to the fact that there is only one CEFA for each string variable and $\varphi$ only cares about the values of registers.

@@

}
%%%%%%%%%%%% the original texts by Denghang %%%%%%%%%
%%%%%%%%%%%% the original texts by Denghang %%%%%%%%%

### 2023-05-05 2023-1-19/sections/cefa.tex
solving NE(CEFA)

The semantics of CEFA is define as follows. Let $\aut = (R, Q, \Sigma, \delta, I, F, \alpha)$ be a CEFA. 
A \emph{configuration} of $\aut$ is a pair $(q, \vec{v})$ where $q \in Q$ and $\vec{v}$ is a vector denoting the values of registers.  
%An \emph{initial configuration} of $\aut$ is $(q_0, \vec{0})$ with $q_0 \in I$, where the value of each register is zero. 
A \emph{run} of $\aut$ on a string $w = a_1 \cdots a_n$ is a sequence $q_0 \xrightarrow[\myvec{v_1}]{a_1} q_1 \cdots q_{n-1}\xrightarrow[\myvec{v_n}]{a_n} q_n$ such that $q_0 \in I$ and $q_{i-1} \xrightarrow[\myvec{v_i}]{a_i} q_i$ for each $i \in [n]$. A run $q_0 \xrightarrow[\myvec{v_1}]{a_1} q_1 \cdots q_{n-1}\xrightarrow[\myvec{v_n}]{a_n} q_n$ is \emph{accepting} if $q_n \in F$ and $\alpha(\myvec{v'}/R)$ is true, where $\myvec{v'} = \sum \limits_{j \in [n]} \myvec{v_j}$. The vector $\myvec{v'}= \sum \limits_{j \in [n]} \myvec{v_j}$ is called the \emph{cost} of an accepting run $q_0 \xrightarrow[\myvec{v_1}]{a_1} q_1 \cdots q_{n-1}\xrightarrow[\myvec{v_n}]{a_n} q_n$. Note here we assume that the initial values of all registers are zero and $\sum \limits_{j \in [n]} \myvec{v_j}$ is the vector of register values after all the transitions in the run are executed.  We use $\myvec{v'} \in \aut(w)$ to denote the fact that there is an accepting run of $\aut$ on $w$ whose cost is $\myvec{v'}$.  
Finally, we define the semantics of a CEFA $\aut$ as $\Lang(\aut) = \{(w; \myvec{v'}) \mid  \myvec{v'} \in \aut(w)\}$.  In particular, if $I \cap F \neq \emptyset$, then $(\varepsilon; \myvec{0}) \in \Lang(\aut)$.

>>>>

The semantics of CEFA is define as follows. Let $\aut = (R, Q, \Sigma, \delta, I, F, \alpha)$ be a CEFA. 
A \emph{configuration} of $\aut$ is a pair $(q, \vec{v})$ where $q \in Q$ and $\vec{v}$ is a vector denoting the values of registers.  
%An \emph{initial configuration} of $\aut$ is $(q_0, \vec{0})$ with $q_0 \in I$, where the value of each register is zero. 
A \emph{run} of $\aut$ on a string $w = a_1 \cdots a_n$ is a sequence $q_0 \xrightarrow[\myvec{v_1}]{a_1} q_1 \cdots q_{n-1}\xrightarrow[\myvec{v_n}]{a_n} q_n$ such that $q_0 \in I$ and $q_{i-1} \xrightarrow[\myvec{v_i}]{a_i} q_i$ for each $i \in [n]$. A run $q_0 \xrightarrow[\myvec{v_1}]{a_1} q_1 \cdots q_{n-1}\xrightarrow[\myvec{v_n}]{a_n} q_n$ is \emph{accepting} if $q_n \in F$ and $\alpha(\myvec{v'}/R)$ is true, where $\myvec{v'} = \sum \limits_{j \in [n]} \myvec{v_j}$. The vector $\myvec{v'}= \sum \limits_{j \in [n]} \myvec{v_j}$ is called the \emph{cost} of an accepting run $q_0 \xrightarrow[\myvec{v_1}]{a_1} q_1 \cdots q_{n-1}\xrightarrow[\myvec{v_n}]{a_n} q_n$. Note here we assume that the initial values of all registers are zero and $\sum \limits_{j \in [n]} \myvec{v_j}$ is the vector of register values after all the transitions in the run are executed.  We use $\myvec{v'} \in \aut(w)$ to denote the fact that there is an accepting run of $\aut$ on $w$ whose cost is $\myvec{v'}$.  
We define the semantics of a CEFA $\aut$, denoted by $\Lang(\aut)$, as $\{(w; \myvec{v'}) \mid  \myvec{v'} \in \aut(w)\}$.  In particular, if $I \cap F \neq \emptyset$, then $(\varepsilon; \myvec{0}) \in \Lang(\aut)$. Moreover, we define the \emph{output} of a CEFA $\aut$, denoted by $\cefaout(\aut)$, as $\{\myvec{v'} \mid  \exists w.\ \myvec{v'} \in \aut(w)\}$.

### 2023-05-05 2023-1-19/sections/preliminaries.tex
solving NE(CEFA)

\paragraph*{Strings and languages.}
A string over $\Sigma$ is a (possibly empty) sequence of elements from $\Sigma$,
denoted by $u, v, w, \ldots$. An empty string is denoted by $\varepsilon$. We use $\Sigma_\varepsilon$ to denote $\Sigma \cup \{\varepsilon\}$. We write $\Sigma^*$ (resp., $\Sigma^+$) for the set of all (resp. nonempty) strings over $\Sigma$.
%Let 
For a string $u$, we use $|u|$ to denote the number of letters in $u$. In particular, $|\varepsilon|=0$. 
%Moreover, for $a \in \Sigma$, let $|u|_a$ denote the number of occurrences of $a$ in $u$. 
Assume $u=a_0\cdots a_{n-1}$ is nonempty and $i<j \in [0,n-1]$. %Then a \emph{position} of $u$ is a number $i \in [|u|]$ (Note that the first position is $1$, instead of  0). In addition, 
We let $u[i]$ denote $a_i$ and $u[i,j]$ for the substring %of $u$ starting from $i$ and ending with $j$ (i.e., 
$a_i\cdots a_j$. 
%\tl{check later for consistency}\zhilin{the indices start from 0, to be consistent with the semantics of substring and indexof}\zhilin{i do use it in section 4}
%
Let $u, v$ be two strings. We use $u \cdot v$ to denote the \emph{concatenation} of $u$ and $v$. A language $L$ over $\Sigma$ is a subset of strings.  
Let $L_1, L_2$ be two languages. Then the concatenation of $L_1$ and $L_2$, denoted by $L_1 \cdot L_2$,  is defined as $\{u \cdot v \mid u \in L_1, v \in L_2\}$. The union (resp. intersection) of $L_1$ and $L_2$, denoted by $L_1 \cup L_2$  (resp. $L_1 \cap L_2$), is defined as $\{u \mid u \in L_1 \mbox{ or } u \in L_2\}$ (resp. $\{u \mid u \in L_1 \mbox{ and } u \in L_2\}$). The complement of $L_1$, denoted by $\overline{L_1}$, is defined as $\{u \mid u \in \Sigma^*, u \not \in L_1\}$. 
The difference of $L_1$ and $L_2$, denoted by $L_1 \setminus L_2$, is defined as $L_1 \cap \overline{L_2}$. 
For a language $L$ and $n \in \Nat$, we define $L^n$ inductively as follows: $L^0= \{\varepsilon\}$, $L^{n+1} = L \cdot L^n$ for every $n \in \Nat$. 
Finally, define $L^* = \bigcup \limits_{n \in \Nat} L^n$.
%, that is, the string $w$ such that $|w|= |u| + |v|$ and for each $i \in [0, |u|-1]$, $w[i]= u[i]$, and for each $i \in [0,|v|-1]$, $w[|u|+i]=v[i]$.

@@

@@

We will also use standard  quantifier-free/existential \emph{linear integer arithmetic} (LIA) formulae, which are typically ranged over by $\phi, \varphi$, etc. 
For a set $X$ of variables, we use $\Phi(X)$ to denote the set of quantifier-free LIA formulas whose free variables are from $X$.
%, which is essentially Presburger

>>>>

\paragraph*{Strings and languages.}
A string over $\Sigma$ is a (possibly empty) sequence of elements from $\Sigma$,
denoted by $u, v, w, \ldots$. An empty string is denoted by $\varepsilon$. We use $\Sigma_\varepsilon$ to denote $\Sigma \cup \{\varepsilon\}$. We write $\Sigma^*$ (resp., $\Sigma^+$) for the set of all (resp. nonempty) strings over $\Sigma$.
%Let 
For a string $u$, we use $|u|$ to denote the number of letters in $u$. In particular, $|\varepsilon|=0$. 
Moreover, for $a \in \Sigma$, let $|u|_a$ denote the number of occurrences of $a$ in $u$. 
Assume $u=a_0\cdots a_{n-1}$ is nonempty and $i<j \in [0,n-1]$. %Then a \emph{position} of $u$ is a number $i \in [|u|]$ (Note that the first position is $1$, instead of  0). In addition, 
We let $u[i]$ denote $a_i$ and $u[i,j]$ for the substring %of $u$ starting from $i$ and ending with $j$ (i.e., 
$a_i\cdots a_j$. 
%\tl{check later for consistency}\zhilin{the indices start from 0, to be consistent with the semantics of substring and indexof}\zhilin{i do use it in section 4}
%
Let $u, v$ be two strings. We use $u \cdot v$ to denote the \emph{concatenation} of $u$ and $v$. A language $L$ over $\Sigma$ is a subset of strings.  
Let $L_1, L_2$ be two languages. Then the concatenation of $L_1$ and $L_2$, denoted by $L_1 \cdot L_2$,  is defined as $\{u \cdot v \mid u \in L_1, v \in L_2\}$. The union (resp. intersection) of $L_1$ and $L_2$, denoted by $L_1 \cup L_2$  (resp. $L_1 \cap L_2$), is defined as $\{u \mid u \in L_1 \mbox{ or } u \in L_2\}$ (resp. $\{u \mid u \in L_1 \mbox{ and } u \in L_2\}$). The complement of $L_1$, denoted by $\overline{L_1}$, is defined as $\{u \mid u \in \Sigma^*, u \not \in L_1\}$. 
The difference of $L_1$ and $L_2$, denoted by $L_1 \setminus L_2$, is defined as $L_1 \cap \overline{L_2}$. 
For a language $L$ and $n \in \Nat$, we define $L^n$ inductively as follows: $L^0= \{\varepsilon\}$, $L^{n+1} = L \cdot L^n$ for every $n \in \Nat$. 
Finally, define $L^* = \bigcup \limits_{n \in \Nat} L^n$.
%, that is, the string $w$ such that $|w|= |u| + |v|$ and for each $i \in [0, |u|-1]$, $w[i]= u[i]$, and for each $i \in [0,|v|-1]$, $w[|u|+i]=v[i]$.

@@

Let $w \in \Sigma^*$. The \emph{Parikh image} of $w$, denoted by $\parikh(w)$, is defined as the function $\eta: \Sigma \rightarrow \Nat$ such that $\eta(a) = |w|_a$ for each $a \in \Sigma$. The \emph{Parikh image} of an NFA $\NFA$, denoted by $\parikh(\NFA)$, is defined as $\{\parikh(w) \mid w \in \Lang(\NFA)\}$.

@@

\paragraph*{Linear integer arithmetic and Parikh images.}
We will also use standard quantifier-free/existential \emph{linear integer arithmetic} (LIA) formulae, which are typically ranged over by $\phi, \varphi$, etc. 
For a set $\mathfrak{X}$ of variables, we use $\Phi(\mathfrak{X})$ to denote the set of quantifier-free LIA formulas whose free variables are from $\mathfrak{X}$. 
For convenience, we use $\varphi(\vec{\anivar})$ with $\myvec{\anivar} = (\anivar_1, \cdots, \anivar_k)$ to denote an LIA formula $\varphi$ whose free variables are from $\{\anivar_1, \cdots, \anivar_k\}$. For an LIA formula $\varphi(\vec{\anivar})$ with $\myvec{\anivar} = (\anivar_1, \cdots, \anivar_k)$ and $\myvec{t} = (t_1, \cdots, t_k)$ be a tuple of integer terms, we use $\varphi[\myvec{t}/\myvec{\anivar}]$ to denote the formula obtained by replacing (simultaneously) $\anivar_i$ with $t_i$ for every $i \in [k]$.
%For convenience, we usually assume that a set $X$ of variables is a vector. 
%, which is essentially Presburger

At last, we recall a result about Parikh images of NFA. 
For each $a \in \Sigma$, let $\anivarz_a$ be an integer variable. Let $\fZ_\Sigma$ denote the set of integer variables $\anivarz_a$ for $a \in \Sigma$. 
Let $\NFA$ be an NFA over the alphabet $\Sigma$. Then 
we say that an LIA formula $\varphi(\fZ_\Sigma)$ defines the Parikh image of $\NFA$, if $\{\eta: \Sigma \rightarrow \Nat \mid \varphi[(\eta(a) /\anivarz_a)_{a \in \Sigma}] \mbox{ is true}\} = \parikh(\NFA)$.

\begin{theorem}[\cite{SSMH04}]\label{thm-nfa-parikh}
Given an NFA $\NFA$,  an existential LIA formula $\varphi_\NFA(\fZ_\Sigma)$ can be computed in linear time that defines the Parikh image of $\NFA$.
\end{theorem}
